---
title: "The Effect of Female Education on Birth Rates in the United States"
author: "Matthew Tillmawitz"
date: "2024-12-09"
output:
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidycensus)
library(tidyverse)
library(magrittr)
library(caret)
library(forcats)
library(kableExtra)
library(ggplot2)
library(dplyr)
library(GGally)
library(psych)
```

```{r load data, message=FALSE, echo=FALSE}
geography <- "public use microdata area"

all_vars <- load_variables(2022, "acs5") # Plain text labels associated with variable codes

education_data <- get_acs( 
  geography = geography,
  table = "B15001",
  year = 2022,
  survey = "acs5",
  cache_table = TRUE
)

birth_data <- get_acs(
  geography = geography,
  table = "B13016",
  year = 2022,
  survey = "acs5",
  cache_table = TRUE
)

# Function to make labels readable
convert_label_to_varname <- function(label) {
  label |>
    str_replace_all("^(Estimate!!|Total:!!)", "") |> # Remove "Estimate!!" and "Total:!!" prefixes
    str_replace_all("!!", " ") |>                    # Replace remaining "!!" with a single space
    str_to_lower() |>                                # Convert to lowercase
    str_replace_all("[^a-z0-9]+", "_") |>            # Replace special characters and spaces with underscores
    str_replace_all("^_|_$", "")                     # Remove leading or trailing underscores
}

# Fetch the labels for the datasets
education_labels <- all_vars |>
  filter(str_detect(name, "B15001")) |>
  dplyr::rename(variable = name) # Rename function was mapping to plyr by default

birth_labels <- all_vars |>
  filter(str_detect(name, "B13016")) |>
  dplyr::rename(variable = name)

# Label the datasets
labeled_education_data <- education_data |>
  left_join(education_labels, by = "variable") |> # Join to get labels into education data
  sapply(convert_label_to_varname) |>             # Convert labels to an easier to understand format
  as_tibble() |>
  filter(str_detect(label, "female")) |>          # Get rid of male data
  select(GEOID, NAME, estimate, moe, label)       # Remove unneeded columns
labeled_education_data %<>% mutate(estimate = as.integer(estimate), moe = as.integer(moe)) # Fix column typing

labeled_birth_data <- birth_labels |>
  left_join(birth_data, by = "variable") |> # Join to get labels into education data
  sapply(convert_label_to_varname) |>       # Convert labels to an easier to understand format
  as_tibble() |>
  arrange(GEOID) |>                         # Reordering for readability when developing
  select(GEOID, NAME, label, estimate, moe) # Keep the required fields
labeled_birth_data %<>% mutate(estimate = as.integer(estimate)) # Fix column typing

# Summarize education data, previously split by age groups
summarised_ed <- labeled_education_data |>
  mutate(prelabel = str_remove(label, r"(\w*_\w*_\d*_\w*_\d*_years_)")) |>   # Removes age range from labels
  mutate(relabel = str_remove(prelabel, r"(\w*_\w*_65_years_\w*_over_)")) |> # 65 and over has a slightly different pattern
  select(!c(label, prelabel)) |>
  group_by(GEOID, NAME, relabel) |>       # relabel is of the form "less_than_9th_grade"
  summarise(pop_estimate = sum(estimate)) # collapse age ranges to education aggregates

# Pivot wider education data, now one row per PUMA
tidied_ed <- summarised_ed |>
  pivot_wider(names_from = relabel, values_from = pop_estimate) |>
  select(!starts_with("total_female_")) # drop the population totals by age

# Truncating birth data labels to remove overly long values
relabeled_birth_data <- labeled_birth_data |> # Label trimming is different for this data
  mutate(trimmed_label = case_when(
    str_detect(label, "total_women_who_had_a_birth_in_the_past_12_months_") ~ paste0("age_", str_extract(label, r"(\d*_to_\d*)"), "_birth"),
    str_detect(label, "total_women_who_did_not_have_a_birth_in_the_past_12_months_") ~ paste0("age_", str_extract(label, r"(\d*_to_\d*)"), "_none"),
    .default = label
  )) |>
  select(GEOID, NAME, trimmed_label, estimate)

# Pivoting birth data, now one row per PUMA
detailed_birth_data <- relabeled_birth_data |>
  pivot_wider(names_from = trimmed_label, values_from = estimate)

# Removing granular columns and calculating birth rate
births_per_thousand <- detailed_birth_data |>
  mutate(births_per_thousand = total_women_who_had_a_birth_in_the_past_12_months / total * 1000) |>
  select(GEOID, NAME, births_per_thousand)

# Join education and fertility data
fertility_and_education <- tidied_ed |>
  left_join(births_per_thousand, by = "GEOID")

# Converting population of education achievement levels to proportions. This is the final dataframe for analysis
prop_fertility_and_education <- fertility_and_education |>
  mutate(
    prop_less_ninth = less_than_9th_grade / total_female,
    prop_no_diploma = `9th_to_12th_grade_no_diploma` / total_female,
    prop_diploma_or_equivalent = high_school_graduate_includes_equivalency / total_female,
    prop_some_college = some_college_no_degree / total_female,
    prop_associate_degree = associate_s_degree / total_female,
    prop_bachelor_degree = bachelor_s_degree / total_female,
    prop_advanced_degree = graduate_or_professional_degree / total_female
   ) |>
  select(GEOID, NAME.x, prop_less_ninth, prop_no_diploma, prop_diploma_or_equivalent, prop_some_college, prop_associate_degree, prop_bachelor_degree, prop_advanced_degree, births_per_thousand) |>
  ungroup()

```

## Project Overview

>- Research Question: Is the pursuit of higher education by women a statistically significant and practically significant negative predictor of birth rates?
>- Dependent Variable: number of births per thousand women in a PUMA.
>- Independent Variables: proportion of the female population who's highest level of academic achievement is the stated level (levels are change through the analysis).
>- Statistical Test: Multiple Linear Regression
>- Data Source: The American Community Survey by the US Census Bureau

```{r, figures-side, fig.show="hide", out.width="50%", message=FALSE}
pop_dist_plot <- labeled_education_data |>
  filter(label == "total_female") |>
  ggplot(aes(x = estimate)) +
  geom_histogram() +
  labs(title = "Distribution of Female Population in Census PUMAs", x = "Female Population Estimate") 
pop_dist_plot

birth_dist_plot <- births_per_thousand |>
  ggplot(aes(x = births_per_thousand)) +
  geom_histogram() +
  labs(title = "Distribution of Birth Rate in Census PUMAs", x = "Births per Thousand Women")
birth_dist_plot
```

## Raw Data {.smaller}

```{r}
head(all_vars) |>
  kbl() |>
  kable_styling()
head(education_data)|>
  kbl() |>
  kable_styling()
```

## Transformed data {.smaller}

```{r}
transformed <- prop_fertility_and_education |>
  select(!NAME.x) |>
  column_to_rownames("GEOID")

head(transformed |> select(prop_less_ninth:prop_some_college)) |>
  kbl() |>
  kable_styling()

head(transformed |> select(prop_associate_degree:births_per_thousand)) |>
  kbl() |>
  kable_styling()
```

```{r female education distribution}
female_pop_totals <- summarised_ed |>
  filter(relabel == "total_female") |>
  pivot_wider(names_from = relabel, values_from = pop_estimate)

percent_education <- summarised_ed |>
  left_join(female_pop_totals, by = "GEOID") |>
  filter(!str_detect(relabel, "total_female")) |>
  mutate(pop_percent = pop_estimate / total_female * 100) |>
  select(GEOID, relabel, pop_percent)

education_order <- c("graduate_or_professional_degree", "bachelor_s_degree", "associate_s_degree", "some_college_no_degree", "high_school_graduate_includes_equivalency",  "9th_to_12th_grade_no_diploma", "less_than_9th_grade")

massaged_ed <- percent_education |>
  mutate(ed_factor = factor(relabel, levels = education_order)) |>
  mutate(renamed = fct_recode(ed_factor, "< 9th Grade" = "less_than_9th_grade", "9th-12th" = "9th_to_12th_grade_no_diploma", "HS diploma or equivalent" = "high_school_graduate_includes_equivalency", "Some College" = "some_college_no_degree", "Associate's Degree" = "associate_s_degree", "Bachelor's Degree" = "bachelor_s_degree", "Graduate Degree or Higher" = "graduate_or_professional_degree")) |>
  select(GEOID, renamed, pop_percent)

ed_dist_plot <- massaged_ed |>
  ggplot(aes(x = renamed, y = pop_percent)) + 
  geom_boxplot() +
  labs(title = "Education Distribution of the Female Population of PUMAs", x = "Highest Level of Education Attained",
       y = "Percent of Female Population") +
  coord_flip()
```

```{r untrans_proportional_education}
raw_model <- lm(births_per_thousand ~ prop_less_ninth + prop_no_diploma + prop_diploma_or_equivalent + prop_some_college + prop_associate_degree + prop_bachelor_degree + prop_advanced_degree, data = prop_fertility_and_education)
```

```{r model residuals, fig.show="hide", out.width="40%", message=FALSE}
ggplot(data = raw_model, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  xlab("Fitted values") +
  ylab("Residuals")

ggplot(data = raw_model, aes(x = .resid)) +
  geom_histogram() +
  xlab("Residuals")

ggplot(data = raw_model, aes(sample = .resid)) +
  stat_qq()
```

```{r transform education, fig.show="hide", out.width="50%"}
trans_df <- prop_fertility_and_education |>
  mutate(
    across(
      .cols = where(is.numeric),
      .fns = ~ predict(BoxCoxTrans(.x), .x),
      .names = "{.col}_bc"
    ),
    .keep = "unused"
  )

education_order_trans <- c("prop_advanced_degree_bc", "prop_bachelor_degree_bc", "prop_associate_degree_bc", "prop_some_college_bc", "prop_diploma_or_equivalent_bc",  "prop_no_diploma_bc", "prop_less_ninth_bc")

trans_vis <- trans_df |>
  select(!c(NAME.x, births_per_thousand_bc)) |>
  pivot_longer(cols = c(prop_less_ninth_bc:prop_advanced_degree_bc), names_to = "ed_level", values_to = "adjusted_proportion") |>
  mutate(ed_factor = factor(ed_level, levels = education_order_trans)) |>
  mutate(renamed = fct_recode(ed_factor, "< 9th Grade" = "prop_less_ninth_bc", "9th-12th" = "prop_no_diploma_bc", "HS diploma or equivalent" = "prop_diploma_or_equivalent_bc", "Some College" = "prop_some_college_bc", "Associate's Degree" = "prop_associate_degree_bc", "Bachelor's Degree" = "prop_bachelor_degree_bc", "Graduate Degree or Higher" = "prop_advanced_degree_bc")) 

trans_vis |>
  filter(ed_level %in% c("prop_less_ninth_bc", "prop_no_diploma_bc", "prop_bachelor_degree_bc", "prop_advanced_degree_bc")) |>
  ggplot(aes(x = renamed, y = adjusted_proportion)) + 
  geom_boxplot() +
  labs(title = "Adjusted Education Distribution of the Female Population of PUMAs", x = "Highest Level of Education Attained",
       y = "Adjusted Measure of Female Population") +
  coord_flip()

trans_vis |>
  filter(ed_level %in% c("prop_diploma_or_equivalent_bc", "prop_some_college_bc", "prop_associate_degree_bc")) |>
  ggplot(aes(x = renamed, y = adjusted_proportion)) + 
  geom_boxplot() +
  labs(title = "Adjusted Education Distribution of the Female Population of PUMAs", x = "Highest Level of Education Attained",
       y = "Adjusted Measure of Female Population") +
  coord_flip()
```

## Adjusted Multiple Linear Regression {.smaller}

```{r proportional_education}
proportional_model <- lm(births_per_thousand_bc ~ prop_less_ninth_bc + prop_no_diploma_bc + prop_diploma_or_equivalent_bc + prop_some_college_bc + prop_associate_degree_bc + prop_bachelor_degree_bc + prop_advanced_degree_bc, data = trans_df)

summary(proportional_model)
```

```{r adjusted residuals, fig.show="hide", out.width="40%", message=FALSE}
ggplot(data = proportional_model, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  xlab("Fitted values") +
  ylab("Residuals")

ggplot(data = proportional_model, aes(x = .resid)) +
  geom_histogram() +
  xlab("Residuals")

ggplot(data = proportional_model, aes(sample = .resid)) +
  stat_qq()
```

```{r milestone regression}
prop_milestone <- prop_fertility_and_education |>
  mutate(prop_any_college = prop_some_college + prop_associate_degree + prop_bachelor_degree,
         prop_no_college = prop_less_ninth + prop_no_diploma + prop_diploma_or_equivalent) |>
  select(GEOID, NAME.x, prop_any_college, prop_no_college, prop_advanced_degree, births_per_thousand) |>
  mutate(
    across(
      .cols = where(is.numeric),
      .fns = ~ predict(BoxCoxTrans(.x), .x),
      .names = "{.col}_bc"
    ),
    .keep = "unused"
  )
milestone_model <- lm(births_per_thousand_bc ~ prop_no_college_bc + prop_any_college_bc + prop_advanced_degree_bc, data = prop_milestone)

# summary(milestone_model)
```

## Milestone Regression {.smaller}

```{r degree regression}
no_cor_degree_model <- lm(births_per_thousand_bc ~ prop_any_college_bc + prop_advanced_degree_bc, data = prop_milestone)

summary(no_cor_degree_model)
```

```{r milestone residuals, fig.show="hide", out.width="40%", message=FALSE}
ggplot(data = no_cor_degree_model, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  xlab("Fitted values") +
  ylab("Residuals")

ggplot(data = no_cor_degree_model, aes(x = .resid)) +
  geom_histogram() +
  xlab("Residuals")

ggplot(data = no_cor_degree_model, aes(sample = .resid)) +
  stat_qq()
```

```{r simple regression}
prop_college <- prop_fertility_and_education |>
  mutate(prop_any_college = prop_associate_degree + prop_bachelor_degree + prop_advanced_degree + prop_some_college) |>
  select(GEOID, NAME.x, prop_any_college, births_per_thousand) |>
  mutate(
    across(
      .cols = where(is.numeric),
      .fns = ~ predict(BoxCoxTrans(.x), .x),
      .names = "{.col}_bc"
    ),
    .keep = "unused"
  )

simple_model <- lm(births_per_thousand_bc ~ prop_any_college_bc, data = prop_college)
```

```{r simple residuals, fig.show="hide", out.width="40%", message=FALSE}
ggplot(data = simple_model, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  xlab("Fitted values") +
  ylab("Residuals")

ggplot(data = simple_model, aes(x = .resid)) +
  geom_histogram() +
  xlab("Residuals")

ggplot(data = simple_model, aes(sample = .resid)) +
  stat_qq()
```

## Conclusion

>- Simply pursuing a higher education is not a statistically significant predictor of birth rates.

>- The proportion of women who have attained an advanced degree held the majority of the predictive power when focusing on the population that pursued higher education.


























